---
title: "Run Targets - Three Lakes System Decision Support System pipeline"
---

# Purpose

This script loads all necessary packages to run the {targets} pipeline for the
Three Lakes System Decision Support System (TLS_DSS) then runs it! This R 
Markdown document is not meant to be "knit", rather to walk the user through
the process of setting up and running this workflow. 


## Prerequsites

There are a number of packages required for this workflow. First, we install 
the necessary packages using a `package_installer()` custom function.

Define package installer function

```{r package_installer}
package_installer <- function(x) {
  if (x %in% installed.packages()) {
    print(paste0("{", x ,"} package is already installed."))
    } else {
      install.packages(x, repos = "http://cran.us.r-project.org")
      print(paste0("{", x ,"} package has been installed."))
    }
  }
```

Check for packages required to run this workflow, install them if not already
present. 

```{r walk_package_installer}
packages <- c("crew",
              "httr2",
              "rvest",
              "tarchetypes",
              "targets",
              "tidyverse",
              "zoo")

# note we use base-r lapply and not map since tidy is not loaded (and possibly
# not installed) yet
lapply(packages, package_installer)
```


## Run the targets pipeline and output a network graph.

This workflow currently uses 6 cores via the crew worker from {targets}. Some
targets are manually set to not use parallel processing using `deployment = "main"`
in the definition of a target. 

This code chunk will inform you of how many cores are available on your computer,
if the number of cores is the same or less than 6, you should change the workers
argument in the _targets.R file ~ line 10 to be fewer than those detected by the
following code chunk:

```{r}
parallel::detectCores()
```

And this code chunk will run the pipeline:

```{r run_targets}
library(targets)

tar_make()
```


### Create a network diagram of the workflow.

```{r see_targets_net}
tar_visnetwork()
```

