---
title: "Data Reduction"
---

Based on the baseline model, we can probably drop some of the features we used
for modeling. Here, we'll drop some of those that we consider redundant based
on the SHAP plots. This script requires that 02 has been run and is still in the
Global environment (it uses the test/val years created there.)

```{r}
if (!exists("test")) {"Run 02_tvt_split before continuing"}
```

To keep:
- water temp at both depths: m1
- solar radiation: today, m1, m2
- air temp: today mean/min/max, m1 mean/min/max
- rel hum: mean m1, m2
- pump: m1, m2, m3
- wind: mean/max m1, today mean/max
- nf: m1-m4
- chip: m1-m4

```{r}
keep_vars <- c("date", "mean_1m_temp_degC", "mean_0_5m_temp_degC",
               "mean_1m_temp_degC_m1", "mean_0_5m_temp_degC_m1",
               "total_solar_radiation", "total_solar_radiation_m1", "total_solar_radiation_m2",
               "mean_air_temp", "min_air_temp", "max_air_temp",
               "mean_air_temp_m1", "min_air_temp_m1", "max_air_temp_m1",
               "mean_rel_hum_m1", "mean_rel_hum_m2", 
               "pump_cfs_m1", "pump_cfs_m2", "pump_cfs_m3",
               "mean_wind", "max_wind", "mean_wind_m1", "max_wind_m1",
               "nf_cfs_m1", "nf_cfs_m2", "nf_cfs_m3", "nf_cfs_m4",
               "chipmunk_cfs_m1", "chipmunk_cfs_m2", "chipmunk_cfs_m3", "chipmunk_cfs_m4")

ar_data <- read_csv("model_submodule/model_dev/data/ar_data.csv")

ar_reduced <- ar_data %>% 
  select(all_of(keep_vars))

write_csv(ar_reduced, "model_submodule/model_dev/data/ar_reduced.csv")
```

Now, let's do this for all the TVT files, too.

```{r}
test_data <- ar_reduced %>% 
  filter(year(date) %in% test, complete.cases(.))
write_csv(test_data, 
          paste0("model_submodule/model_dev/operational_model/data/test.csv"))


walk2(.x =  list(val_1, val_2, val_3, val_4),
      .y = list("_1", "_2", "_3", "_4"),
     .f = ~ {
       val_data <- ar_reduced %>% 
         filter(year(date) %in% .x, complete.cases(.))
       write_csv(val_data, 
                 paste0("model_submodule/model_dev/operational_model/data/validation",
                        .y,
                        ".csv"))
       train_data <- reduce(list(ar_reduced, test_data, val_data),
                            anti_join) %>% 
         filter(complete.cases(.))
       write_csv(train_data, 
                 paste0("model_submodule/model_dev/operational_model/data/training",
                        .y,
                        ".csv"))
     })
```


