---
title: "Calculating Performance of the Baseline"
---

This script calculates performance of a baseline as a way to gauge performance of the arNN. For this application, we will use a persistence model, that is, assuming yesterday-is-today.

```{r}
library(tidyverse)
library(Metrics)

data <- read_csv("model_submodule/model_dev/data/SMR_buoy_daily_summary.csv") %>% 
  filter(year(date) %in% c(2015,2022))
data$mean_1m_temp_degC_p1 = lag(data$mean_1m_temp_degC, 1)
data$mean_0_5m_temp_degC_p1 = lag(data$mean_0_5m_temp_degC, 1) 

reduced <- data %>% 
  select(mean_1m_temp_degC, mean_1m_temp_degC_p1, mean_0_5m_temp_degC, mean_0_5m_temp_degC_p1) 

reduced <- reduced[complete.cases(reduced),]
```

```{r}
mse(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
rmse(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
mae(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
mape(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
```

```{r}
mse(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
rmse(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
mae(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
mape(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
```

And do the same at a 7-day time horizon

```{r}
reduced$mean_1m_temp_degC_p7 = lag(reduced$mean_1m_temp_degC, 7)
reduced$mean_0_5m_temp_degC_p7 = lag(reduced$mean_0_5m_temp_degC, 7)

reduced <- reduced[complete.cases(reduced),]
```

```{r}
mse(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p7)
rmse(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p7)
mae(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p7)
mape(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p7)
```

```{r}
mse(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p7)
rmse(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p7)
mae(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p7)
mape(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p7)
```

Now compare that with 7-day forecast from the arNN. These data were created in the operational model/forecast_walkthrough_control... file

```{r}
forecast_control_2015 <- read_csv("model_submodule/model_dev/operational_model/test_forecast/forecasted_temp_2015_control_collated.csv")
forecast_control_2022 <- read_csv("model_submodule/model_dev/operational_model/test_forecast/forecasted_temp_2022_control_collated.csv")
```

We need the 7th day of each of these forecasts
```{r}
all_forecasts <- bind_rows(forecast_control_2015, forecast_control_2022) 
unique_fore_date <- unique(all_forecasts$forecast_date)

seventh_day <- map(.x = unique_fore_date, 
                   .f = ~ {
                     all_forecasts %>% 
                       filter(forecast_date == .x, valid_date == .x + days(6)) %>% 
                       summarize(ens_mean_1m_temp_degC = mean(mean_1m_temp_degC),
                                 ens_mean_0_5m_temp_degC = mean(mean_0_5m_temp_degC),
                                 .by = valid_date)
                   }) %>% 
  bind_rows() %>% 
  rename(date = valid_date) 

data_fore <- inner_join(data, seventh_day)
```

```{r}
mse(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
rmse(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
mae(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
mape(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
```

```{r}
mse(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
rmse(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
mae(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
mape(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
```

Now compare that with 7-day forecast from the arNN. These data were created in the operational model/forecast_walkthrough_control... file

```{r}
forecast_control_2015 <- read_csv("model_submodule/model_dev/tiny_model/test_forecast/forecasted_temp_2015_control_collated.csv")
forecast_control_2022 <- read_csv("model_submodule/model_dev/tiny_model/test_forecast/forecasted_temp_2022_control_collated.csv")
```

We need the 7th day of each of these forecasts
```{r}
all_forecasts <- bind_rows(forecast_control_2015, forecast_control_2022) 
unique_fore_date <- unique(all_forecasts$forecast_date)

seventh_day <- map(.x = unique_fore_date, 
                   .f = ~ {
                     all_forecasts %>% 
                       filter(forecast_date == .x, valid_date == .x + days(6)) %>% 
                       summarize(ens_mean_1m_temp_degC = mean(mean_1m_temp_degC),
                                 ens_mean_0_5m_temp_degC = mean(mean_0_5m_temp_degC),
                                 .by = valid_date)
                   }) %>% 
  bind_rows() %>% 
  rename(date = valid_date) 

data_fore <- inner_join(data, seventh_day)
```

```{r}
mse(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
rmse(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
mae(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
mape(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
```

```{r}
mse(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
rmse(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
mae(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
mape(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
```