---
title: "Calculating Performance of the Baseline"
---

This script calculates performance of a baseline as a way to gauge performance of the arNN. For this application, we will use a persistence model, that is, assuming yesterday-is-today.

```{r}
library(tidyverse)
library(Metrics)

data <- read_csv("model_submodule/model_dev/data/SMR_buoy_daily_summary.csv") 

# just grab the test set
data_2015 <- data %>% filter(year(date) == 2015)
data_2022 <- data %>% filter(year(date) == 2022)
# make sure all dates are included in the set
dates_2015 <- tibble(date = seq.Date(ymd("2015-01-01"), ymd("2015-12-31"), by = "day"))
dates_2022 <- tibble(date = seq.Date(ymd("2022-01-01"), ymd("2022-12-31"), by = "day"))
# merge!
data_2015 <- full_join(dates_2015, data_2015)
data_2022 <- full_join(dates_2022, data_2022)

# calculate the lagged temp (by 1 day)
data_2015$mean_1m_temp_degC_p1 = lag(data_2015$mean_1m_temp_degC, 1)
data_2015$mean_0_5m_temp_degC_p1 = lag(data_2015$mean_0_5m_temp_degC, 1) 
data_2022$mean_1m_temp_degC_p1 = lag(data_2022$mean_1m_temp_degC, 1)
data_2022$mean_0_5m_temp_degC_p1 = lag(data_2022$mean_0_5m_temp_degC, 1) 

reduced <- full_join(data_2015, data_2022) %>% 
  select(mean_1m_temp_degC, mean_1m_temp_degC_p1, mean_0_5m_temp_degC, mean_0_5m_temp_degC_p1) %>% 
  filter(complete.cases(.))

```

```{r}
mse(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
rmse(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
mae(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
mape(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
```

> mse(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
[1] 0.1900516
> rmse(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
[1] 0.4359491
> mae(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
[1] 0.3393106
> mape(reduced$mean_1m_temp_degC, reduced$mean_1m_temp_degC_p1)
[1] 0.02212511

```{r}
mse(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
rmse(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
mae(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
mape(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
```

> mse(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
[1] 0.1082525
> rmse(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
[1] 0.3290175
> mae(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
[1] 0.2611108
> mape(reduced$mean_0_5m_temp_degC, reduced$mean_0_5m_temp_degC_p1)
[1] 0.01925052
And do the same at a 7-day time horizon

```{r}
data_2015$mean_1m_temp_degC_p7 = lag(data_2015$mean_1m_temp_degC, 7)
data_2015$mean_0_5m_temp_degC_p7 = lag(data_2015$mean_0_5m_temp_degC, 7)
data_2022$mean_1m_temp_degC_p7 = lag(data_2022$mean_1m_temp_degC, 7)
data_2022$mean_0_5m_temp_degC_p7 = lag(data_2022$mean_0_5m_temp_degC, 7)

reduced_7 <- full_join(data_2015, data_2022) %>% 
  select(mean_1m_temp_degC, mean_1m_temp_degC_p7, mean_0_5m_temp_degC, mean_0_5m_temp_degC_p7) %>% 
  filter(complete.cases(.))
```

```{r}
mse(reduced_7$mean_1m_temp_degC, reduced_7$mean_1m_temp_degC_p7)
rmse(reduced_7$mean_1m_temp_degC, reduced_7$mean_1m_temp_degC_p7)
mae(reduced_7$mean_1m_temp_degC, reduced_7$mean_1m_temp_degC_p7)
mape(reduced_7$mean_1m_temp_degC, reduced_7$mean_1m_temp_degC_p7)
```

> mse(reduced_7$mean_1m_temp_degC, reduced_7$mean_1m_temp_degC_p7)
[1] 1.396092
> rmse(reduced_7$mean_1m_temp_degC, reduced_7$mean_1m_temp_degC_p7)
[1] 1.181563
> mae(reduced_7$mean_1m_temp_degC, reduced_7$mean_1m_temp_degC_p7)
[1] 0.9470429
> mape(reduced_7$mean_1m_temp_degC, reduced_7$mean_1m_temp_degC_p7)
[1] 0.06141142

```{r}
mse(reduced_7$mean_0_5m_temp_degC, reduced_7$mean_0_5m_temp_degC_p7)
rmse(reduced_7$mean_0_5m_temp_degC, reduced_7$mean_0_5m_temp_degC_p7)
mae(reduced_7$mean_0_5m_temp_degC, reduced_7$mean_0_5m_temp_degC_p7)
mape(reduced_7$mean_0_5m_temp_degC, reduced_7$mean_0_5m_temp_degC_p7)
```

> mse(reduced_7$mean_0_5m_temp_degC, reduced_7$mean_0_5m_temp_degC_p7)
[1] 0.8331743
> rmse(reduced_7$mean_0_5m_temp_degC, reduced_7$mean_0_5m_temp_degC_p7)
[1] 0.9127838
> mae(reduced_7$mean_0_5m_temp_degC, reduced_7$mean_0_5m_temp_degC_p7)
[1] 0.7377067
> mape(reduced_7$mean_0_5m_temp_degC, reduced_7$mean_0_5m_temp_degC_p7)
[1] 0.05337264

Now compare that with 7-day forecast from the arNN. These data were created in the operational model/forecast_walkthrough_control... file

```{r}
forecast_control_2015 <- read_csv("model_submodule/model_dev/operational_model/test_forecast/forecasted_temp_2015_control_collated.csv")
forecast_control_2022 <- read_csv("model_submodule/model_dev/operational_model/test_forecast/forecasted_temp_2022_control_collated.csv")
```

We need the 7th day of each of these forecasts
```{r}
all_forecasts <- bind_rows(forecast_control_2015, forecast_control_2022) 
unique_fore_date <- unique(all_forecasts$forecast_date)

seventh_day <- map(.x = unique_fore_date, 
                   .f = ~ {
                     all_forecasts %>% 
                       filter(forecast_date == .x, valid_date == .x + days(6)) %>% 
                       summarize(ens_mean_1m_temp_degC = mean(mean_1m_temp_degC),
                                 ens_mean_0_5m_temp_degC = mean(mean_0_5m_temp_degC),
                                 .by = valid_date)
                   }) %>% 
  bind_rows() %>% 
  rename(date = valid_date) 

data_fore <- inner_join(data, seventh_day)
```

```{r}
mse(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
rmse(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
mae(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
mape(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
```

> mse(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
[1] 0.5505593
> rmse(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
[1] 0.7419968
> mae(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
[1] 0.620197
> mape(data_fore$ens_mean_1m_temp_degC, data_fore$mean_1m_temp_degC)
[1] 0.04122693
```{r}
mse(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
rmse(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
mae(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
mape(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
```

> mse(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
[1] 0.4090926
> rmse(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
[1] 0.6396035
> mae(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
[1] 0.4962873
> mape(data_fore$ens_mean_0_5m_temp_degC, data_fore$mean_0_5m_temp_degC)
[1] 0.03637066