---
title: "Data Exploration and Manipulation for NN Application"
editor_options: 
  markdown: 
    wrap: 80
---

```{r}
library(tidyverse)
```

# Purpose

This markdown walks through data exploration of the buoy data, flow data, and
weather data for feature creation, feature reduction, and applies normalization
and/or standardization to features and targets prior to neural network (NN)
model development and application.

-   Feature: any variable used as an input to a model, sometimes used
    interchangeably with variable or parameter, but specifically refers to the
    inputs of a machine learning model
-   Target: the variable predicted by a model, sometimes called output
-   Feature creation: the act of aggregating existing variables into a feature
    that could be used for modeling; e.g. creating a daily mean from 3-h data,
    creating categorical variables/features from zero-inflated variables
-   Feature reduction: the act of reducing features to remove redundancy and
    decrease the tendency towards overtraining of the NN
-   Normalization/standardization: When using a NN it is imperative to have
    values for the features and targets that are small (as in between -1 to 1).
    While this is not always, it does help the model converge and can reduce
    over-reliance on large values/features with large ranges in prediction.

# Buoy Data

Quick visualization of the buoy data that will become our target variable:

```{r}
buoy <- read_csv("model_submodule/model_dev/data/SMR_buoy_daily_summary.csv")
buoy_vert <- buoy %>% 
  pivot_longer(cols = !date,
               names_to = "parameter",
               values_to = "value") %>% 
  mutate(depth = case_when(grepl("1m", parameter) ~ "near surface",
                           grepl("0_5m", parameter) ~ "integrated",
                           .default = NA_character_),
         type = case_when(grepl("min", parameter) ~ "minium daily value",
                          grepl("mean", parameter) ~ "average daily value",
                          grepl("max", parameter) ~ "maximum daily value",
                          grepl("sd", parameter) ~ "standard deviation of the mean",
                          .default = NA_character_))
buoy_vert %>% 
  filter(!grepl("sd", parameter)) %>% 
  ggplot(., aes(x = date, y = value, color = type)) +
  geom_point() +
  facet_grid(depth ~ ., scales = "free_y") +
  theme_bw() +
  scale_color_viridis_d() +
  theme(legend.position = "bottom")
```

Nothing much to report home about here. We really just care about the mean for
each of these depths at this point, so let's create a target object from the
buoy data with just the mean of the near-surface and integrated data.

```{r}
target <- buoy %>% 
  select(date, mean_1m_temp_degC, mean_0_5m_temp_degC)

# and quick histograms
hist(target$mean_1m_temp_degC)
hist(target$mean_0_5m_temp_degC)
```

# Weather Data
