---
title: "Data Exploration and Manipulation for NN Application"
editor_options: 
  markdown: 
    wrap: 80
---

```{r}
library(tidyverse)
library(imputeTS)
```

# Purpose

This markdown walks through data exploration of the buoy data, flow data, and
weather data for feature creation, feature reduction, and applies normalization
and/or standardization to features and targets prior to neural network (NN)
model development and application.

-   Feature: any variable used as an input to a model, sometimes used
    interchangeably with variable or parameter, but specifically refers to the
    inputs of a machine learning model
-   Target: the variable predicted by a model, sometimes called output
-   Feature creation: the act of aggregating existing variables into a feature
    that could be used for modeling; e.g. creating a daily mean from 3-h data,
    creating categorical variables/features from zero-inflated variables
-   Feature reduction: the act of reducing features to remove redundancy and
    decrease the tendency towards overtraining of the NN
-   Normalization/standardization: When using a NN it is imperative to have
    values for the features and targets that are small (as in between -1 to 1,
    or z-score standardization using the mean and standard deviation). While
    this is not always necessary, it does help the model converge and can reduce
    over-reliance on large values/features with large ranges in prediction.

# Buoy Data

Quick visualization of the 10 years of buoy data that will become our target
variable:

```{r}
buoy <- read_csv("model_submodule/model_dev/data/SMR_buoy_daily_summary.csv")
buoy_vert <- buoy %>% 
  pivot_longer(cols = !date,
               names_to = "parameter",
               values_to = "value") %>% 
  mutate(depth = case_when(grepl("1m", parameter) ~ "near surface",
                           grepl("0_5m", parameter) ~ "integrated",
                           .default = NA_character_),
         type = case_when(grepl("min", parameter) ~ "minium daily value",
                          grepl("mean", parameter) ~ "average daily value",
                          grepl("max", parameter) ~ "maximum daily value",
                          grepl("sd", parameter) ~ "standard deviation of the mean",
                          .default = NA_character_))
buoy_vert %>% 
  filter(!grepl("sd", parameter)) %>% 
  ggplot(., aes(x = date, y = value, color = type)) +
  geom_point() +
  facet_grid(depth ~ ., scales = "free_y") +
  theme_bw() +
  scale_color_viridis_d() +
  theme(legend.position = "bottom")
```

Nothing much to report home about here. We really just care about the mean for
each of these depths at this point, so let's create a target object from the
buoy data with just the mean of the near-surface and integrated data.

```{r}
target <- buoy %>% 
  select(date, mean_1m_temp_degC, mean_0_5m_temp_degC)

# and quick histograms
hist(target$mean_1m_temp_degC)
hist(target$mean_0_5m_temp_degC)
```

And these data will need to be standardized for use in the NN, we'll use a
z-score for these values.

**z = (x-μ)/σ**

```{r}
standardize_function <- function(values_to_be_standardized) {
  mean_value <- mean(values_to_be_standardized, na.rm = T)
  sd_value <- sd(values_to_be_standardized, na.rm = T)
  standardized_values <- (values_to_be_standardized - mean_value)/sd_value
  # and we'll return mean, sd, and the standardized values, naming them for
  # easier use later
  list("mean" = mean_value, "sd" = sd_value, "values" = standardized_values)
}

# calculate z scores
z_1m <- standardize_function(target$mean_1m_temp_degC)
z_int <- standardize_function(target$mean_0_5m_temp_degC)

target <- target %>% 
  mutate(z_1m = z_1m$values,
         z_int = z_int$values) %>% 
  select(-c(mean_1m_temp_degC, mean_0_5m_temp_degC))
```

# Weather Data

These data have been processed to "look like" the NOAA GEFS forecast data. We'll
summarize these as we will the GEFS data so that we have equivalent data between
the two sources. We know that surface water temperature will be driven by air
temperature and solar radiation, and that wind contributes to mixing within the
waterbody when it's not strongly stratified (and can generally affect the
strength of stratification otherwise) and that periods of high temperature and
low wind are times when algal biomass proliferate.

```{r}
met <- read_csv("model_submodule/model_dev/data/SMR_met_three_hour_summary.csv") %>% 
  arrange(date, fcast)

met_summary <- met %>% 
  # don't use na.rm = T here, because we want complete 3-h "forecasts" only
  summarize(mean_air_temp = mean(air_temp),
            max_air_temp = max(air_temp),
            min_air_temp = min(air_temp), 
            mean_wind = mean(ave_wind),
            max_wind = max(ave_wind),
            min_wind = min(ave_wind),
            mean_rel_hum = mean(rel_hum),
            max_rel_hum = max(rel_hum),
            total_solar_radiation = sum(solar_rad),
            .by = date) 

met_summary %>% 
  pivot_longer(cols = mean_air_temp:total_solar_radiation)  %>% 
  mutate(parameter_group = case_when(grepl("temp", name) ~ "Air Temperature",
                                     grepl("wind", name) ~ "Wind Speed",
                                     name == "total_solar_radiation" ~ "Total Solar Radiation",
         grepl("hum", name) ~ "Relative Humidity")) %>% 
  ggplot(., aes(x = date, y = value)) +
  geom_point() +
  facet_grid(parameter_group ~ ., scales = "free_y") +
  theme_bw()

```

Okay, some data gaps in temperature 2019-2020 that appear to be due to missing
data twice per day. It's probably safe to say we can impute those linearly
between neighboring three-hour periods. This is not a perfect method, but it
might help us retain some data from the record that would otherwise be lost.

```{r}
# look at where the missing data is using {imputeTS}
met %>% 
  pull(air_temp) %>% 
  ggplot_na_distribution(.)

met %>% 
  pull(ave_wind) %>% 
  ggplot_na_distribution(.)

met %>% 
  pull(rel_hum) %>% 
  ggplot_na_distribution()

met %>% 
  pull(solar_rad) %>% 
  ggplot_na_distribution(.)

```

Looks like we can just do this for temp, relative humidity, and solar radiation.
Temperature first.

```{r}
# now we'll impute using a simple spline regression imputation technique using {imputeTS}
interp_linear <- met %>% arrange(date, fcast) %>% pull(air_temp) %>% na_interpolation(option = "linear", maxgap = 2)
interp_spline <- met %>% arrange(date, fcast) %>% pull(air_temp) %>% na_interpolation(option = "spline", maxgap = 2)

met$temp_linear <- interp_linear
met$temp_spline <- interp_spline

met %>% 
  filter(between(year(date), 2019, 2020)) %>% 
  ggplot(., aes(x = date, y = air_temp)) +
  geom_point() +
  geom_point(aes(y = temp_linear), color = "orange") +
  geom_point(aes(y = temp_spline), color = "green") +
  theme_bw()

met %>% 
  filter(between(date, ymd("2020-06-01"), ymd("2020-06-15"))) %>% 
  mutate(datetime = case_when(fcast == "f000" ~ ymd_hm(paste(date, "00:00")),
                              fcast == "f003" ~ ymd_hm(paste(date, "03:00")),
                              fcast == "f006" ~ ymd_hm(paste(date, "06:00")),
                              fcast == "f009" ~ ymd_hm(paste(date, "09:00")),
                              fcast == "f012" ~ ymd_hm(paste(date, "12:00")),
                              fcast == "f015" ~ ymd_hm(paste(date, "15:00")),
                              fcast == "f018" ~ ymd_hm(paste(date, "18:00")),
                              fcast == "f021" ~ ymd_hm(paste(date, "21:00")))) %>% 
  ggplot(., aes(x = datetime, y = air_temp)) +
  geom_point() +
  geom_line(aes(y = temp_linear), color = "orange") +
  geom_line(aes(y = temp_spline), color = "green") +
  theme_bw()
```

Spline looks better here, so let's go with that. I assume that will be the case
for solar radiation and relative humidity, too, since it's a diurnal pattern, so
just go ahead and apply to the dataset since there are pretty few instances of
missing data.

```{r}
met$rad_spline <- met %>% arrange(date, fcast) %>% pull(solar_rad) %>% na_interpolation(option = "spline", maxgap = 2)
met$hum_spline <- met %>% arrange(date, fcast) %>% pull(rel_hum) %>% na_interpolation(option = "spline", maxgap = 2)
```

And now summarize and re-plot.

```{r}
met_summary <- met %>% 
  # don't use na.rm = T here, because we want complete 3-h "forecasts" only
  summarize(mean_air_temp = mean(temp_spline),
            max_air_temp = max(temp_spline),
            min_air_temp = min(temp_spline), 
            mean_wind = mean(ave_wind),
            max_wind = max(ave_wind),
            min_wind = min(ave_wind),
            mean_rel_hum = mean(hum_spline),
            max_rel_hum = max(hum_spline),
            total_solar_radiation = sum(rad_spline),
            .by = date) 

met_summary %>% 
  pivot_longer(cols = mean_air_temp:total_solar_radiation)  %>% 
  mutate(parameter_group = case_when(grepl("temp", name) ~ "Air Temperature",
                                     grepl("wind", name) ~ "Wind Speed",
                                     name == "total_solar_radiation" ~ "Total Solar Radiation",
                                     grepl("hum", name) ~ "Relative Humidity")) %>% 
  ggplot(., aes(x = date, y = value)) +
  geom_point() +
  facet_grid(parameter_group ~ ., scales = "free_y") +
  theme_bw()
```

Just a small gap in temp in early 2020, which shouldn't affect our ability to
implement the NN.

And now, we'll do some quick correlations and determine if we'll keep all the
summary values

```{r}
plot(met_summary)
cor(met_summary$mean_air_temp, met_summary$max_air_temp, use = "complete.obs")
cor(met_summary$mean_air_temp, met_summary$min_air_temp, use = "complete.obs")
cor(met_summary$max_air_temp, met_summary$min_air_temp, use = "complete.obs")
```

Ridiculously strong relationship between mean/max air temp, but we know that
both of these parameters actually can "tell us" about what is happening. For the
purposes of this model, we are not going to remove any variables at this time.
And now we can standardize these, but we don't need all the data, just data
generally the data around the buoy deployment.

```{r}
buoy %>% 
  mutate(yr = year(date)) %>% 
  summarize(min_date = min(date), 
            max_date = max(date), 
            .by = yr)
```

Let's filter for May 1 through end of Oct 31.

```{r}
met_reduced <- met_summary %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 5, 10)) %>% 
  select(-month)
```

And now we can standardize across the parameters we care about and create a new
dataframe with that data.

```{r}
params <- names(met_reduced)[!grepl("date", names(met_reduced))]
param_list <- map(.x = params, .f = ~ met_reduced %>% pull(!!.x)) 
names(param_list) <- params
met_z <- map(.x = param_list, .f = ~standardize_function(.x))

# and grab the info we want to store
met_z_values <- map(.x = met_z, .f = ~ .x$values) %>% 
  bind_cols() %>% 
  mutate(date = met_reduced$date) %>% 
  relocate(date)
```

# Flow Data

And now we'll look at the flow data.

```{r}
flow <- read_csv("model_submodule/model_dev/data/flow_daily.csv") %>% 
  filter(year(date) < 2024)

flow %>%
  pivot_longer(cols = pump_cfs:chipmunk_cfs) %>% 
  ggplot(., aes(x = date, y = value)) +
  geom_point() +
  facet_grid(name ~ ., scales = "free_y") +
  theme_bw()
```

And let's look at data gaps

```{r}
# look at where the missing data is using {imputeTS}
flow %>% 
  pull(pump_cfs) %>% 
  ggplot_na_distribution(.)

flow %>% 
  pull(nf_cfs) %>% 
  ggplot_na_distribution(.)

flow %>% 
  pull(ni_cfs) %>% 
  ggplot_na_distribution(.)

flow %>% 
  pull(adams_cfs) %>% 
  ggplot_na_distribution(.)

flow %>% 
  pull(chipmunk_cfs) %>% 
  ggplot_na_distribution(.)

```

A lot of the missing values are over the winter, which is totally okay. For
these, I think we should do a linear interpolation to fill in the short \< 2 day
gaps.

```{r}
flow_interp <- flow %>% 
  mutate(across(.cols = pump_cfs:chipmunk_cfs,
                .fns = ~ na_interpolation(., option = "linear", maxgap = 2)))

# look at where the missing data is using {imputeTS}
flow_interp %>% 
  pull(pump_cfs) %>% 
  ggplot_na_distribution(.)

flow_interp %>% 
  pull(nf_cfs) %>% 
  ggplot_na_distribution(.)

flow_interp %>% 
  pull(ni_cfs) %>% 
  ggplot_na_distribution(.)

flow_interp %>% 
  pull(adams_cfs) %>% 
  ggplot_na_distribution(.)

flow_interp %>% 
  pull(chipmunk_cfs) %>% 
  ggplot_na_distribution(.)
```

Great. Let's do a quick correlation plot.

```{r}
plot(flow_interp)
cor(flow_interp$ei_cfs, flow_interp$ni_cfs, use = "complete.obs")
```

Pretty high correlation between East and North inlet to Grand, and Chipmunk kind
of tells the tale otherwise. To reduce the need for excessive water balance
calculations/applications, we will drop EI, NI and Adams for modeling purposes
since most of that data is embedded in the Chipmunk flow data. We'll also filter
out data for the same time period as the met data.

```{r}
flow_reduced <- flow_interp %>%
  select(-c(ni_cfs, ei_cfs, adams_cfs)) %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 5, 10)) %>% 
  select(-month)
```

And now to standardize.

```{r}
params <- names(flow_reduced)[!grepl("date", names(flow_reduced))]
param_list <- map(.x = params, .f = ~ flow_reduced %>% pull(!!.x)) 
names(param_list) <- params
flow_z <- map(.x = param_list, .f = ~standardize_function(.x))

# and grab the info we want to store
flow_z_values <- map(.x = flow_z, .f = ~ .x$values) %>% 
  bind_cols() %>% 
  mutate(date = flow_reduced$date) %>% 
  relocate(date)
```

# Collate and add previous days

Now, we'll bring everything together. For the most part this is pretty
straightforward - we want all of our data in a single row to play nice as a
table. We want tomorrows temperature as the target, but are parameter will all
be previous day's observations/measurements except met, which we can provide
tomorrow's data as well (this is the equivalent of NOAA GEFS data).

## Create parameters of previous days

We are using an autoregressive neural network for this application, meaning we
will give the model yesterday (and a few previous day's temperature. For the
most part, a good model within aquatic science is that which performs better
than persistence ("yesterday-is-today"), so we're going to embed that into the
model. For the purposes of this model, we're going to provide the previous 3
days of water temperature. A super simple and quick way to do this is to adjust
the 'date' column and join together.

```{r}
params <- names(target)[!grepl("date", names(target))]

temp_m1_3 <- map(.x = 1:3, function(i) {
  target %>%
    mutate(date = date + days(i)) %>%
    rename_with(.cols = params,
                .fn = ~ paste0(.x, "_m", i))
  }) %>% 
  reduce(., full_join)

ar_temp <- full_join(target, temp_m1_3)
```

Now for met. Similar idea here, but we're going to embed a bit more air
temperature memory (10d), since that has a seasonal signal.

```{r}
params <- names(met_z_values)[!grepl("date", names(met_z_values))]

met_m1_3 <- map(.x = 1:3, function(i) {
  met_z_values %>%
    mutate(date = date + days(i)) %>%
    rename_with(.cols = params,
                .fn = ~ paste0(.x, "_m", i))
  }) %>% 
  reduce(., full_join)

air_temp_m4_7 <- met_z_values %>% 
  select(date, contains("air_temp"))
air_temp_m4_7 <- map(.x = 4:7, function(i) {
  air_temp_m4_7 %>%
    mutate(date = date + days(i)) %>%
    rename_with(.cols = contains("air_temp"),
                .fn = ~ paste0(.x, "_m", i))
  }) %>% 
  reduce(., full_join)

# join "tomorrow's" met with previoius day summaries.
ar_met <- reduce(list(met_z_values, met_m1_3, air_temp_m4_7),
                 full_join)
```

And now flow data. We're going to do previous 7 here. But we also want to use
operational pumping of "tomorrow" as a parameter.

```{r}
params <- names(flow_z_values)[!grepl("date", names(flow_z_values))]

flow_m1_7 <- map(.x = 1:7, function(i) {
  flow_z_values %>%
    mutate(date = date + days(i)) %>%
    rename_with(.cols = params,
                .fn = ~ paste0(.x, "_m", i))
  }) %>% 
  reduce(., full_join)

ar_flow <- flow_z_values %>% 
  select(date, pump_cfs) %>% 
  full_join(., flow_m1_7)
```

## Collate all, reduce to complete observations.

```{r}
ar_data <- reduce(list(ar_temp, ar_met, ar_flow),
                  full_join)
```

And there we go! There are a lot of variables in here right now, but we'll worry
about that later, when we create a final model, where we'll use just a
proportion of these that are consistently meaningful across all ensemble models.

```{r}
write_csv(ar_data, "model_submodule/model_dev/data/ar_data.csv")
```
