---
title: "Train Validate Test Split"
---

Here, we're just creating splits across the years to create an ensemble of models.
Unlik in model dev, we will NOT have a test set for this data, as we want
to use as much data as possible to inform the operational deployment.

```{r}
library(tidyverse)
ar_data <- read_csv("model_submodule/model_dev/data/ar_data.csv")

# get list of unique years
years <- unique(year(ar_data$date))

```

Let's reduce the variables here to those that are in the tiny model.

```{r}
keep_cols <- c("date", "mean_1m_temp_degC", "mean_0_5m_temp_degC",
               "mean_1m_temp_degC_m1", "mean_0_5m_temp_degC_m1",
               "total_solar_radiation", "total_solar_radiation_m1",
               "mean_air_temp", "min_air_temp", "max_air_temp",
               "mean_air_temp_m1", "min_air_temp_m1", "max_air_temp_m1",
               "mean_rel_hum_m1", 
               "pump_cfs_m1", 
               "mean_wind", "max_wind", "mean_wind_m1", "max_wind_m1",
               "nf_cfs_m1",
               "chipmunk_cfs_m1")


ar_data <- ar_data %>% 
  select(all_of(keep_cols))
```


We're going to randomly choose 2 years for the test set, and then randomly pair
the remaining years for validation sets (and those not in the list are the training), 
resulting in 4 datasets to create models that will be the ensemble.

```{r}
set.seed(47)

val_1 <- sample(years, size = 2, replace = F)
val_2 <- sample(years[!years %in% c(val_1)], size = 2, replace = F)
val_3 <- sample(years[!years %in% c(val_1, val_2)], size = 2, replace = F)
val_4 <- sample(years[!years %in% c(val_1, val_2, val_3)], size = 2, replace = F)
val_5 <- years[!years %in% c(val_1, val_2, val_3, val_4)]
```

And now, make a list of the training sets and save files

```{r}
walk2(.x =  list(val_1, val_2, val_3, val_4, val_5),
      .y = list("_1", "_2", "_3", "_4", "_5"),
     .f = ~ {
       val_data <- ar_data %>% 
         filter(year(date) %in% .x, complete.cases(.))
       write_csv(val_data, 
                 paste0("model_submodule/tiny_model_operation/data/validation",
                        .y,
                        ".csv"))
       train_data <- reduce(list(ar_data, val_data),
                            anti_join) %>% 
         filter(complete.cases(.))
       write_csv(train_data, 
                 paste0("model_submodule/tiny_model_operation/data/training",
                        .y,
                        ".csv"))
     }
)

```

